---
title: "BAYESIAN MODELING OF MICROBIOME DATA FOR DIFFERENTIAL ABUNDANCE ANALYSIS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tutorial

The following script is used to perform differential abundance analysis for microbiome count data using the proposed Zero-inflated Negative Binomial model with Dirichlet Process Prior (ZINB-DPP) model in the manuscript Q.Li _et al. 2019_ https://arxiv.org/abs/1902.08741.

## Loading Data

We first load a simulated $n$-by-$p$ micribiome count dataset. Here, we simulate $n = 24$ and $p = 1000$.

```{r get_Y}
Y = read.table("data/simulation_single.txt")
Y = as.matrix(Y)
n = nrow(Y); p = ncol(Y)
```

We also need the phenotype indicator vector of the $n$ samples. Each element in the indicator vector is a binary variable with value of either $1$ or $0$. Here, we have 2 groups with a balanced group size of $12$.

```{r get_z}
group_info = rownames(Y)
z_vector = substr(group_info, nchar(group_info), nchar(group_info))
z_vector = as.numeric(z_vector) - 1
```

Before fitting the model, we get the true discriminating taxa by creating a binary vector `gamma_true`, where $1$ represents that the taxon is discriminating and $0$ otherwise.

```{r get_true_gamma}
gamma_true = rep(0, dim(Y)[2])
gamma_true[grep(".TP", colnames(Y))] = 1
```

## Loading Main Function

We load the main function of the ZINB-DPP model:

```{r load_main_function,message=FALSE }
library(Rcpp)
sourceCpp('utility/core_zinb_x5.cpp')
```

We set all the model parameter by default and fit the model:

```{r fit_zinbdpp, cache = TRUE}
s_sizefactor = rep(1, n)
Niter = 10000  
S = matrix(1, 1, 1)
G = matrix(1, 1, 1)
b = 1
h = 10

# set a random seed #
set.seed(198642)

# run zinb-dpp model #
zinbdpp_output = zinb_model_estimator(Y = Y, z = z_vector,
s = s_sizefactor, iter = Niter, DPP = TRUE, b = b, h = h,S = S, G = G,store = TRUE, aggregate = FALSE, MRF = F)

```

## Evaluation

Now, we evaluate our model by visualizing the result of the differential abundance analysis.

```{r load_lbr, include=FALSE, message=FALSE}
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())
```

```{r viz_zinbdpp, message=FALSE}
# load function to calculate the threshold that 
# controls the Bayesian false discovery rate 
source('utility/user_functions.R');

zinbdpp_PPI = zinbdpp_output$gamma_ppi
c_gamma = BayFDR(PPI = zinbdpp_PPI, alpha = 0.05)
ppi_df = data.frame(x = seq(1, length(zinbdpp_PPI)),
                    y = zinbdpp_PPI)
ppi_point_df = data.frame(x = which(gamma_true == 1),
                          y = zinbdpp_PPI[which(gamma_true == 1)])
ggplot(ppi_df, 
       aes(x = x,xend = x, y = 0, yend = y) ) + 
  geom_segment() +
  geom_point(data = ppi_point_df, aes(x, y), color = 'red')+
  geom_hline(yintercept = c_gamma, linetype = "dashed", color = 'red') + 
  xlab("Taxon index") + ylab("Posterior probability of inclusion")
```


Each vertial line is the posterior proability of inclusion (PPI) of a taxon. The dashed line represent the threshold that control the Bayesian false discovery rate to be less than $0.05$. Taxa whose PPIs exceed the dashed line would be selected as the differentially abundant taxa.

Notice that the red dots are true discriminating taxa. According to the model fitting, we have 

- Number of true positive = `r {sum(zinbdpp_PPI > c_gamma & gamma_true == 1)}`

- Number of false positive = `r {sum(zinbdpp_PPI > c_gamma & gamma_true == 0)}`

- Number of true negative = `r {sum(zinbdpp_PPI <= c_gamma & gamma_true == 0)}`

- Number of false negative = `r {sum(zinbdpp_PPI <= c_gamma & gamma_true == 1)}`


## Size Factor Estimation 

We can also check the estimation of the size factors for all $n = 24$ samples in the data.

```{r sf}
# Load true size factors
s_true = read.table("data/true_size_factor.txt")
s_true = as.vector(s_true)

# Load estimated size factors
s_mcmc = zinbdpp_output$s
si_df_dpp = data.frame(st = s_true, 
                       dpp = apply(s_mcmc, 2, mean),
                       low = apply(s_mcmc, 2, quantile, 0.025),
                       up = apply(s_mcmc, 2, quantile, 0.975))
colnames(si_df_dpp)[1] = "st"

# Compare true and estimated size factors
ggplot(si_df_dpp, aes(st, dpp)) +
  geom_errorbar(aes(ymin = low,ymax = up),width = 0.2)+
  geom_point(colour = "black", size = 2.5, shape=1) +
  ylim(0, 5)+xlim(0, 5)+xlab("True")+ylab("Estimated")+ggtitle("DPP") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")

```

In the figure above, each vertical bar is a $95\%$ credible interval estimation of the corresponding size factor, with the posterior mean in the middle.

## Result Reproduce
Please refer to R code `demo_code.R` to reproduce all the above results.

## Contact 

Qiwei Li, Department of Mathematical Science, The University of Texas at Dallas, Richardson, Texas <liqiwei2000@gmail.com>
