---
title: "BAYESIAN MODELING OF MICROBIOME DATA FOR DIFFERENTIAL ABUNDANCE ANALYSIS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tutorial

The following script is used to perform differential abundance analysis for microbiome count data using the proposed ZINB-DPP model in the manuscript Q.Li _et al. 2019_ https://arxiv.org/abs/1902.08741.

## Loading Data

We first load a simulated $n$-by-$p$ micribiome count dataset. Here, we simulate $n = 24$ and $p = 1000$.

```{r get_Y}
Y = read.table("data/simulation_single.txt")
Y = as.matrix(Y)
n = nrow(Y); p = ncol(Y)
```

We also need the phenotype indicator vector of the $n$ samples. Each element in the indicator vector should be a binary variable with value of either $1$ or $0$.

```{r get_z}
group_info = rownames(Y)
z_vector = substr(group_info, nchar(group_info), nchar(group_info))
z_vector = as.numeric(z_vector) - 1
```

Before fitting the model, we get the true discriminating taxa by creating a binary vector `gamma_true`, where $1$ represents that the taxon is discriminating and $0$ otherwise.

```{r get_true_gamma}
gamma_true = rep(0, dim(Y)[2])
gamma_true[grep(".TP", colnames(Y))] = 1
```

## Loading Main Function

We load the main function of the ZINB-DPP model"

```{r load_main_function,message=FALSE }
library(Rcpp)
sourceCpp('utility/core_zinb_x5.cpp')
```

We set all the model parameter by default and fit the model:

```{r fit_zinbdpp}
s_sizefactor = rep(1, n)
Niter = 10000  
S = matrix(1, 1, 1)
G = matrix(1, 1, 1)
b = 1
h = 10

# set a random seed #
set.seed(198642)

# run zinb-dpp model #
zinbdpp_output = zinb_model_estimator(Y = Y, z = z_vector,
s = s_sizefactor, iter = Niter, DPP = TRUE, b = b, h = h,S = S, G = G,store = TRUE, aggregate = FALSE, MRF = F)

```

## Evaludation 

Now, we evaluate our model by visualizing the result of the differential abundance analysis.

```{r viz_zinbdpp, echo = F,message=FALSE}
source('utility/user_functions.R');
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

s_true = read.table("data/true_size_factor.txt")
s_true = as.vector(s_true)

zinbdpp_PPI = zinbdpp_output$gamma_ppi
c_gamma = BayFDR(PPI = zinbdpp_PPI, alpha = 0.05)
ppi_df = data.frame(x = seq(1, length(zinbdpp_PPI)),
                    y = zinbdpp_PPI)
ppi_point_df = data.frame(x = which(gamma_true == 1),
                          y = zinbdpp_PPI[which(gamma_true == 1)])
ggplot(ppi_df, 
       aes(x = x,xend = x, y = 0, yend = y) ) + 
  geom_segment() +
  geom_point(data = ppi_point_df, aes(x, y), color = 'red')+
  geom_hline(yintercept = c_gamma, linetype = "dashed", color = 'red') + 
  xlab("Feature index") + ylab("Marginal PPI")
```


Each vertial line is the posterior proability of inclusion (PPI) of a taxon. The dashed line represent the threshold that control the Bayesian false discovery rate to be less than $0.05$. Taxa whose PPIs exceed the dashed line would be selected as the differentially abundant taxa.

Notice that the red dots are true discriminating taxa. According to the model fitting, we have 

- Number of true positive = `r {sum(zinbdpp_PPI > c_gamma & gamma_true == 1)}`

- Number of false positive = `r {sum(zinbdpp_PPI > c_gamma & gamma_true == 0)}`

- Number of true negative = `r {sum(zinbdpp_PPI <= c_gamma & gamma_true == 0)}`

- Number of false negative = `r {sum(zinbdpp_PPI <= c_gamma & gamma_true == 1)}`


## Size Factor Estimation 

We can also check the estimation of the size factors for all $n = 24$ samples in the data.

```{r echo = F}
s_mcmc = zinbdpp_output$s
si_df_dpp = data.frame(st = s_true, 
                       dpp = apply(s_mcmc, 2, mean),
                       low = apply(s_mcmc, 2, quantile, 0.025),
                       up = apply(s_mcmc, 2, quantile, 0.975))
colnames(si_df_dpp)[1] = "st"

ggplot(si_df_dpp, aes(st, dpp)) +
  geom_errorbar(aes(ymin = low,ymax = up),width = 0.2)+
  geom_point(colour = "black", size = 2.5, shape=1) +
  ylim(0, 5)+xlim(0, 5)+xlab("True")+ylab("Estimated")+ggtitle("DPP") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")

```

In the figure above, each vertical bar is the $95\%$ credible interval estimation of the size factor, with the posterior mean in the middle.
